Type Classes for Mathematics in Type Theory

### 2010-06-30 jar.tex
Complete outline in introduction.

\begin{document}
\title{Type Classes for Mathematics in Type Theory}
\author{Bas Spitters \and Eelis van der Weegen} % For MSCS \\ Radboud University Nijmegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
The introduction of first-class type classes in the Coq system calls for re-examination of the basic interfaces we use for mathematical formalization in type theory.
% especially since existing solutions fall short from a proof engineering perspective.
We present a new set of foundational type classes for mathematics that take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible. Thus, we address both traditional proof engineering challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.

@@

% ideally should be more or less readable by Haskell person. that would be neat. so that's what the preliminaries should prepare for.

\paragraph{Outline}
In section~\ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes). Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures. In section~\ref{predicateclasses} we show how type classes can make the use of ``unbundled'' purely predicate based interfaces for abstract structures practical. Next, in section~\ref{classes}, we discuss our algebraic hierarchy implemented using such predicate classes.

@@

In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transferred
to any type theory based proof assistant supporting type classes such as
Matita~\cite{asperti2007user}.

@@

Unfortunately, there is actually one annoying wrinkle here, which will also explain why, as the observant reader will have noticed, we do not register these two fields as instance resolution hints (by declaring them with \lstinline|:>|). What we really want these fields to express is ``\emph{if} in a certain context we know something to be a \lstinline|Monoid_Morphism|, \emph{then} realize that the source and target are \lstinline|Monoid|s''. However, the current instance resolution implementation has little support for this style of \emph{forward} reasoning, and is really primarily oriented on \emph{backward} reasoning: had we registered \lstinline|monmor_to| and \lstinline|monmor_from| as instance resolution hints, we would in fact be saying ``\emph{if} trying to establish that something is a \lstinline|Monoid|, \emph{then} try finding a \lstinline|Monoid_Morphism| to or from it'', which quickly degenerates into a wild goose chase. We will return to this point in section~\ref{conclusion}. % todo: do this

@@

\section{Numerical interfaces}

@@

For more complex properties such as injectivity of partially applied addition and multiplication, we take a longer detour. As part of our universal algebra theory, we have proved that proofs of statements in the language of an equational theory can be transferred between isomorphic implementations. Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us tranfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings. We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in Section~\ref{quote}.

@@

\section{Quoting with type classes}\label{quote}

@@

For the present example, we define an ad-hoc term language for monoids
\begin{lstlisting}
Inductive Expr (V: Type) := Mult (a b: Expr V) | One | Var (v: V).
\end{lstlisting}
The expression type is parametrized over the set of variable indices. Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variables, thereby introducing bound problems and a need for dummy variables.

@@

\section{Sequences and universes}

@@

\section{Conclusions}\label{conclusion}

>>>>

\begin{document}
\title{Type Classes for Mathematics in Type Theory}
\author{Bas Spitters \and Eelis van der Weegen} % For MSCS \\ Radboud University Nijmegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
The introduction of first-class type classes in the Coq system calls for re-examination of the basic interfaces used for mathematical formalization in type theory.
% especially since existing solutions fall short from a proof engineering perspective.
We present a new set of foundational type classes for mathematics and take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible. Thus, we address both traditional proof engineering challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.

@@

% ideally should be more or less readable by Haskell person. that would be neat. so that's what the preliminaries should prepare for.
In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transferred
to any type theory based proof assistant supporting type classes such as
Matita~\cite{asperti2007user}.

\paragraph{Outline}
In section~\ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes). Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures. In section~\ref{predicateclasses} we show how type classes can make the use of ``unbundled'' purely predicate based interfaces for abstract structures practical. Next, in section~\ref{classes}, we discuss our algebraic hierarchy implemented using such predicate classes. In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers. In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}. In section~\ref{sequences} we hint at an interface for sequences but describe how a limitation in the Coq system makes its use problematic. We end with conclusions in section~\ref{conclusions}.

@@

@@

Unfortunately, there is actually one annoying wrinkle here, which will also explain why, as the observant reader will have noticed, we do not register these two fields as instance resolution hints (by declaring them with \lstinline|:>|). What we really want these fields to express is ``\emph{if} in a certain context we know something to be a \lstinline|Monoid_Morphism|, \emph{then} realize that the source and target are \lstinline|Monoid|s''. However, the current instance resolution implementation has little support for this style of \emph{forward} reasoning, and is really primarily oriented on \emph{backward} reasoning: had we registered \lstinline|monmor_to| and \lstinline|monmor_from| as instance resolution hints, we would in fact be saying ``\emph{if} trying to establish that something is a \lstinline|Monoid|, \emph{then} try finding a \lstinline|Monoid_Morphism| to or from it'', which quickly degenerates into a wild goose chase. We will return to this point in section~\ref{conclusions}.

@@

\section{Numerical interfaces}\label{numbers}

@@

For more complex properties such as injectivity of partially applied addition and multiplication, we take a longer detour. As part of our universal algebra theory, we have proved that proofs of statements in the language of an equational theory can be transferred between isomorphic implementations. Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us tranfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings. We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in Section~\ref{quoting}.

@@

\section{Quoting with type classes}\label{quoting}

@@

For the present example, we define an ad-hoc term language for monoids
\begin{lstlisting}
Inductive Expr (V: Type) := Mult (a b: Expr V) | One | Var (v: V).
\end{lstlisting}
The expression type is parameterized over the set of variable indices. Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variables, thereby introducing bound problems and a need for dummy variables.

@@

\section{Sequences and universes}\label{sequences}

@@

\section{Conclusions}\label{conclusions}
